---
title: "Example panel performance"
author: "Aimee R Taylor"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{paneljudge_example}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: ref.bib
csl: vancouver.csl
---

# Introduction 

In this vignette, to illustrate the functionality of the R package **paneljudge**, we inspect the performance of an example panel using example marker allele frequencies based on *Plasmodium falciparum* data from Colombia. Example frequencies are distributed with the package; data used to estimate them are not.   

First we provide a summary of the example panel in terms of its marker count, and its markers' positions, diversities, effective cardinalities and cardinalities. These quantities provide a rough idea of how informative the panel is regarding relatedness inference @taylor2019. Second, we simulate and fit data under the hidden Markov model (HMM) described in @taylor2019. The second step provides a more direct summary of how informative the panel is regarding relatedness inference. 

At present, we do not consider model specification; we do not account for uncertainty around input allele frequency estimates; we do not consider relatedness between pairs of haploid genotypes simulated using different allele frequencies; we do not account for marker drop-out (markers that fail to produce useful data, e.g. because they a monomorphic). Otherwise stated, at present, the performance of a panel is judged in its most favourable light; it will likely perform less well in reality.

To view all the source code for this vignette, including that used to generate the plots, simply type `edit(vignette("paneljudge_example"))`. 

## Glossary

In this vignette, a "marker" refers to a genetic polymorphism that can be typed in order to characterise population genetic variation, while "allele" is used to refer to a realisation of a marker for a particular individual. Examples of markers include single nucleotide polmorphisms (SNPs), microsatellites, and microhaplotypes. The latter refers to a region of the genome that encompasses two or more SNPs and is small enough to be typed using an amplicon (a chemical by-product of genotyping). Examples of alleles include a nucleotide base at a SNP, an amino acid at a non-synonymous SNP, a repeat length at a microsatellite and a sequence of nucleotide bases at a microhaplotype. Arguably, it would be more intuitive to use microhaplotype to refer to an allele at a marker that is a genome region that encompasses a SNP multiplet, and to use amplicon to refer to said marker. However, an amplicon is a chemical by-product of genotyping (not a biological entity), and can be used to type single SNPs as well as SNP multiplets. As such, and for consistency with @taylor2019, microhaplotype is used to refer to the marker. 

We use "markers" and "alleles" in order to be as general as possible. The HMM on which package **paneljudge** relies does not care if markers are SNPs, microsatellites or microhaplotypes as long as they can be modelled as categorical random variables whose realisations (i.e. alleles) are unordered. To this end, microsatellites and microhaplotypes are treated as point polymorphisms and their alleles are enumerated. Enumeration serves only to label the realisations. As such, the ordinal nature of microsatellite alleles is not accounted for even if repeat lengths are enumerated in increasing order. 



```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>", 
  echo = FALSE,
  fig.width=7, fig.height=4, fig.fullwidth = TRUE 
)
default_par <- par() # Save default par to pass to on.exit() at end
```

```{r setup}
library(paneljudge) # Load and attach 
rm(list = ls()) # For cleanliness 
set.seed(1) # For reproducibility
```

# Marker summaries

For relatedness inference, panels with many, evenly spaced and highly diverse markers are informative @taylor2019.  
```{r marker count}
m <- nrow(markers) # Number of markers
num_chr <- length(unique(markers$chrom)) # Number of chromosomes
max_chr <- max(markers$chrom) # Max. number of chromosomes
av_marker_length <- mean(markers$length) # Average marker lengtah
sd_marker_length <- sd(markers$length) # Standard deviation of marker length

# Do any marker starts preceed the stop of the previous marker? 
overlap <- (sapply(unique(markers$chrom), function(chr){ # For each chromosome
  ind <- markers$chrom == chr # chromosom indicators 
  any(markers$Start[ind][-1] - markers$Stop[ind][-sum(ind)] < 0) 
}))

# Single out CTSA from the GTseq panel
CTSA = markers$Amplicon_name[grepl("z", markers$Amplicon_name)]
CTSA_ind = which(markers$Amplicon_name %in% CTSA)
```

The example panel has `r m` markers distributed across `r num_chr` of `r max_chr` chromosomes: 

```{r position plot}
# Amplicon position plot
par(mar = c(5.1,4.1,1,1))
plot(markers$pos, markers$chrom, pch = "|", las = 1,
     ylab = 'Chromosome', xlab = 'Chromosomal position (base pair)')

# Add chromosome lengths
for(chr in unique(markers$chrom)){
  segments(x0 = 0, x1 = chr_lengths[chr], y0 = chr, y1 = chr)
}

# Add CTSA annotations from GTseq panel
text(x = markers$pos[CTSA_ind], 
     y = markers$chrom[CTSA_ind], 
     labels = gsub('z', '', CTSA), 
     pos = 3, offset = 0.5, cex = 0.5)
```

The markers in the example panel are microhaplotypes: short regions of the genome (whose length can be spanned by a single amplicon) that contain two of more SNPs. The positions of four of the microhaplotypes (TRAP, AMA1, CSP and SERA2) are annotated. Although the microhaplotypes are polymorphisms with physical length, we plot them as points (using their mid-points), since their lengths are barely visible on the above scale due to their small size (average `r round(av_marker_length)` base pairs; standard deviation: `r round(sd_marker_length)` base pairs). Some markers are very close together (minimum distance between marker mid-points `r min(markers$distances)` base pairs). However, none overlap: 

```{r overlap}
writeLines(sprintf("Do any marker starts preceed the stop of the previous marker? %s", any(overlap)))
```

```{r marker length}
par(mfrow = c(1,2), mar = c(5.1,4.1,1,1))
hist(markers$length, col = 'gray', main = ' ', las = 1,  
     xlab = 'Marker length', ylab = 'Marker count')
hist(markers$distances, col = 'gray', main = ' ', las = 1, 
     xlab = 'Distance between marker mid-points', ylab = 'Marker count')
```

Next we compute marker diversity, effective cardinality, cardinality and maximum potential diversity. Using **paneljudge** functions, marker diversities and effective cardinalities were calculated from example marker allele frequencies without accounting for the fact that they are estimates (e.g. without correcting for finite sample sizes or considering uncertainty). Marker cardinality is either a theoretical maximum (e.g. $2^{3}$ for a microhaplotype spanning three biallelic single nucleotide polymorphisms), or an observed count. Here, the observed count is calculated (i.e. the per-marker count of alleles with non-zero frequency). Maximum diversity is a function of cardinality assuming equifrequent alleles. 

```{r, echo = T}
# Compute marker summaries using paneljudge functions 
diversities <- compute_diversities(frequencies$Colombia)
eff_cardinalities <- compute_eff_cardinalities(frequencies$Colombia)

# Per-marker allele count:
cardinalities <- apply(frequencies$Colombia, 1, function(x){sum(x > 0)}) 

# Compute max. diversity as a function of cardinality 
max_diversities <- sapply(1:max(cardinalities), function(x){1-sum(rep(1/x, x)^2)})
```

Marker diversity is the probability of picking two different alleles per marker. For an outbred diploid, it is equal to heterozygosity. Highly diverse markers are markers with many equifrequent alleles. For relatedness inference, the potential informativeness of a marker thus scales with its allele count (its cardinality). Alleles are rarely equifrequent, however. Effective cardinality is an allele count that accounts for inequifrequent alleles: it is equal to cardinality if all alleles are equifrequent; otherwise, it is less. Across the `r m` markers of the example panel, marker diversity and effective cardinality range from `r paste(round(range(diversities),2), collapse = ' to ')` and `r paste(round(range(eff_cardinalities),2), collapse = ' to ')`, respectively:

```{r position plot coloured, fig.height=10}
par(mfrow = c(2,1), mar = c(5.1,4.1,1,1))

# Transform markers summaries to [0,1] for plotting
eff_cardinalities_01 <- (eff_cardinalities - min(eff_cardinalities)) / diff(range(eff_cardinalities))
diversities_01 <- (diversities - min(diversities)) / diff(range(diversities))

# Function to map transformed marker summaries onto colour
col_mapper <- function(num){
  y <- colorRamp(c('blue','yellow','red'))(num)
  z <- rgb(y[,1], y[,2], y[,3], maxColorValue = 255)
  return(z)
}

# Marker position coloured by diversity
plot(markers$pos, markers$chrom, pch = "|", las = 1,
     ylab = 'Chromosome', xlab = 'Chromosomal position (base pair)',
     col = col_mapper(diversities_01))
# Add chromosome lengths
for(chr in unique(markers$chrom)){
  segments(x0 = 0, x1 = chr_lengths[chr], y0 = chr, y1 = chr)
}
legend('bottomright', bty = 'n',
       pch = '_', col = col_mapper(seq(1,0,length.out = m)),
       legend = round(c(max(diversities), rep(NA, m-2), min(diversities)),2),
       inset = 0.05, y.intersp = 0.05)
text(x = 2687330, y = 4.7, labels = 'Diversity', srt = 90)

# Marker position coloured by cardinality
plot(markers$pos, markers$chrom, pch = "|", las = 1,
     ylab = 'Chromosome', xlab = 'Chromosomal position (base pair)',
     col = col_mapper(eff_cardinalities_01))
# Add chromosome lengths
for(chr in unique(markers$chrom)){
  segments(x0 = 0, x1 = chr_lengths[chr],y0 = chr, y1 = chr)
}
legend('bottomright', bty = 'n',
       pch = '_', col = col_mapper(seq(1,0,length.out = m)),
       legend = round(c(max(eff_cardinalities), rep(NA, m-2), min(eff_cardinalities)),2),
       inset = 0.05, y.intersp = 0.05)
text(x = 2687330, y = 4.7, labels = 'Effective cardinality', srt = 90)
```


Both marker diversity and effective cardinality are measures of marker informativeness regarding relatedness inference. Personally, I find effective cardinality more intuitive because potential effective cardinality scales linearly with cardinality, whereas potential diversity does not. The average effective cardinality multiplied by marker count, `r round(mean(eff_cardinalities) * m)`, provides a rough summary of the informativeness of the entire panel. 


```{r}
par(mfrow = c(1,2), mar = c(5.1,4.1,1,1))

# Summary plots of marker summaries
plot(NULL, bty = 'n', las = 1,
     ylim = range(cardinalities), xlim = range(cardinalities),
     ylab = 'Effective cardinality (effective allele count)',
     xlab = 'Cardinality (observed allele count)')
polygon(x = c(1:max(cardinalities), max(cardinalities):1),
        y = c(rep(1, max(cardinalities)), max(cardinalities):1),
        col = 'gray', border = NA)
points(x = cardinalities, y = eff_cardinalities, pch = "_")
lines(x = 1:max(cardinalities), y = 1:max(cardinalities), pch = "_", col = 'blue')
legend('topleft', lty = 1, col = c('black', 'blue'), inset = 0.01,
       legend = c('Observed', 'Maximum poss. given cardinality'),
       cex = 0.5)

plot(NULL, bty = 'n', yaxt = 'n',
     ylim = c(0, 1), xlim = range(cardinalities),
     ylab = 'Diversity', xlab = 'Cardinality (observed allele count)')
polygon(x = c(1:max(cardinalities), max(cardinalities):1),
        y = c(max_diversities, rep(0, max(cardinalities))),
        col = 'gray', border = NA)

axis(side = 2, at = max_diversities[c(1,2,3,4,5,12)], las = 1,
     labels = format(max_diversities[c(1,2,3,4,5,12)], digits = 2, drop0trailing = F),
     cex.axis = 0.75, tck = -0.02)
# axis(side = 2, at = max_diversities[seq(1,12,2)], las = 1,
#      labels = rep('', length(seq(1,12,2))),
#      cex.axis = 0.4, tck = 0.02, adj = 0.09)
points(x = cardinalities, y = diversities, pch = "_")
lines(x = 1:max(cardinalities), y = max_diversities, pch = "_", col = 'blue')
legend('topleft', lty = 1, col = c('black', 'blue'), inset = 0.01,
       legend = c('Observed', 'Maximum poss. given cardinality'),
       cex = 0.5)
```

# Simulated data  

Using the example panel and marker allele frequencies, for various data-generating values of relatedness, $r$, we first simulate some data on a pair of haploid genotypes; second estimate $r$ (and a switch rate parameter, $k$, whose data-generating value is fixed in this vignette; see footnote); and third compute 95\% confidence intervals (CIs) around the estimates $\hat{r}$ and $\hat{k}$:

```{r, echo = T}
#=============================================================
# Simulate n genotype pairs for various r and fixed k
#=============================================================
# Data-generating relatedness values (named s.t. mle_CIs are named)
rs <- c("0.01"=0.01, "0.25"=0.25, "0.50"=0.50, "0.75"=0.75, "0.99"=0.99)
k <- 5 # Data-generating switch rate parameter value
n <- 5 # Number of pairs to per simulate per r in rs
fs <- frequencies$Colombia # example marker allele frequencies
ds <- markers$distances # distances between marker mid-points

mle_CIs <- lapply(rs, function(r) {
  sapply(1:n, function(i) {

    # First simulate genotype pair
    Ys <- simulate_Ys(fs, ds, k, r)

    # Second, estimate r and k
    krhat <- estimate_r_and_k(fs, ds, Ys)

    # Third, compute confidence intervals (CIs)
    CIs <- compute_r_and_k_CIs(fs, ds, khat = krhat['khat'], rhat = krhat['rhat'])

    # End of function
    return(c(krhat['rhat'], CIs['rhat',]))
  })
})
```

For each data-generating value of relatedness we simulated `r n` haploid genotype pairs. The 95\% CIs around the relatedness estimates show that, despite considerable uncertainty, the panel is informative regarding relatedness inference across a range of data-generating values (an uninformative panel would have 95\% CIs that span the entire 0 to 1 range): 

```{r}
# Plot CIs
par(mfrow = c(1, length(rs)), mar = c(0,0,0,0), oma = c(4,5,1,1))
for(r in names(mle_CIs)){
  rCIs <- mle_CIs[[as.character(r)]]
  rhat_order <- sort.int(rCIs['rhat',], index.return = T)$ix
  plot(NULL, ylim = c(0,1), xlim = c(1,n),
       ylab = '', xlab = '', yaxt = 'n')
  polygon(x = c(1:n, n:1),
          y = c(rCIs['2.5%', rhat_order], rev(rCIs['97.5%', rhat_order])),
          col = 'gray', border = NA)
  points(y = rCIs['rhat', rhat_order], x = 1:n, col = 'black', pch = 20)
  abline(h = as.numeric(r))
}
mtext(side = 1, outer = T,
      text = expression("Sample pair index per data-generating"~italic(r)), line = 3)
axis(side = 2, outer = T, at = rs, las = 1)
mtext(side = 2, outer = T, text = expression(hat(italic(r))), line = 3)
legend('center', lwd = 1, legend = expression("data-generating"~italic(r)), bty = 'n', cex = 0.75)
```

The variation in $\hat{r}$ around the data-generating $r$ (black horizontal line) is partly due to limited panel informativeness and partly due to the finite length of the genome, i.e. Mendelian sampling @hill2011. Whole genome simulation is required to partition these two sources of variation; see footnote below.

Instead of plotting the CIs directly, we can plot the CI widths as a indicator of informativeness (an uninformative CI having width one). Since `r n` haploid genotype pairs were simulated per data-generating value of $r$, we can also compute and plot the root mean square error (RMSE) as an measure of informativeness.

```{r, echo = T}
CI_widths <- sapply(mle_CIs, function(x){ # Calculate CI widths
  x["97.5%",] - x["2.5%",]
})

RMSEs <- sapply(names(mle_CIs), function(r){ # Calculate RMSEs 
  sqrt(mean((mle_CIs[[r]] - as.numeric(r))^2))
})
```


```{r}
par(mfrow = c(1,2), mar = c(5.1,4.1,1,1), cex.lab = 0.75)

# Plot CI lengths (box plots or violin plots)
boxplot(x = CI_widths,
        ylim = c(0,0.5),
        xlab = expression("Data-generating"~italic(r)),
        ylab = "95% confidence interval width")

# Plot RMSEs
plot(y = RMSEs, x = rs, type = 'b', pch = 20,
     ylim = c(0,0.5), xlim = c(0,1),
     xlab = expression("Data-generating"~italic(r)),
     ylab = expression("Root mean square error given data-generating"~italic(r)))
```

## Footnote: Mendelian sampling 

Relatedness, $r$, is defined as a probability of identity-by-descent (IBD) averaged over the genome @taylor2019. It differs to realised relatedness, which is the fraction of the genome that is IBD, due to the finite length of the genome (i.e. due to Mendelian sampling @hill2011): for a genome of finite length many realised relatednesses are compatible with a given probability of IBD, meaning that there will always be some variance in realised relatedness around $r$. Consequently, the RMSE of $\hat{r}$ compared to the data-generating $r$ (and the width of a CI computed using the parametric bootstrap) asymptotes to some small but non-zero value when more and more markers are typed @taylor2019. 

Considering a panel with a finite marker count, variation in $\hat{r}$ around the data-generating $r$ is partly due to the informativeness of the panel and partly due to Mendelian sampling @hill2011. To set-aside variation due to Mendelian sampling and thus focus entirely on panel performance, we could compare $\hat{r}$ to realised relatedness (e.g. when we compute RMSE). However, this is computationally expensive as it requires whole genome simulation:

1. simulate a haploid genotype pair across the entire genome given a data-generating value of $r$;
2. compute realised relatedness by averaging over latent IBD states; 
3. infer relatedness using only data simulated at the panel markers;
4. compare $\hat{r}$ to realised relatedness. 

## Footnote: switch rate parameter

The switch rate parameter, $k$, controls the rate at which the Markov chain switches between the latent states (IBD and not IBD) in the HMM @taylor2019. When considering data on a limited number of markers in a panel (versus whole genome sequence data), we treat the switch rate as a nuisance parameter (since it has little effect on the estimate, $\hat{r}$, yet is near-on impossible to estimate precisely without whole genome sequence data), focusing instead on the relatedness parameter $r$.


# References

<!-- see https://www.zotero.org/styles for citation styles -->
